---
title: "Code book"
author: "Marcos Marvá"
date: "05/22/2015"
output: pdf_document
---


## Overview on the experiment, variables and data
In this section we reproduce <http://archive.ics.uci.edu/ml/datasets/Human+Activity+Recognition+Using+Smartphones> in order to briefly describe the experiments that yielded the data. 
The experiments have been carried out with a group of 30 volunteers within an age bracket of 19-48 years. Each person performed six activities 

  * walking
  * walking_upstairs, 
  * walking_downstairs, 
  * sitting, 
  * standing, 
  * laying

wearing a smartphone (Samsung Galaxy S II) on the waist. Using its embedded accelerometer and gyroscope, the authors captured 3-axial linear acceleration and 3-axial angular velocity at a constant rate of 50Hz. The obtained dataset has been randomly partitioned into two sets, where 70% of the volunteers was selected for generating the training data and 30% the test data.

The sensor signals (accelerometer and gyroscope) were pre-processed by applying noise filters and then sampled in fixed-width sliding windows of 2.56 sec and 50% overlap (128 readings/window). The sensor acceleration signal, which has gravitational and body motion components, was separated using a Butterworth low-pass filter into body acceleration and gravity. The gravitational force is assumed to have only low frequency components, therefore a filter with 0.3 Hz cutoff frequency was used. From each window, a vector of features was obtained by calculating variables from the time and frequency domain.


### Brief description of variables and data 
The data can be downloaded from <https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip>.

As we have already said, the data was splitted into two subgroups  labeled as "training" data and "test" data. 

The files 

  * subject_test.txt
  * subject_train.txt
  
assingn a number between 1 and 30 to the individual that performed each experiment and belong to the test and train subset, respectively. Each individual performed each activity more than once. 

 The files 

  * y_test.txt
  * y_train.txt
  
assingn a number between 1 and 6 to the activity performed by each individual. 
The files 

  * X_test.txt
  * X_train.txt

contain the values of the 561 variables (we also refer to them as _features_) for each individual belonging to the test and train subset,respectively. 

### Feature selection 
The features selected for this database come from the accelerometer 3-axial raw signals tAcc-XYZ and the gyroscope 3-axial raw signals tGyro-XYZ. These time domain signals (prefix 't' to denote time) were captured at a constant rate of 50 Hz. Then they were filtered using a median filter and a 3rd order low pass Butterworth filter with a corner frequency of 20 Hz to remove noise. 

Similarly, the acceleration signal was then separated into body acceleration signals (tBodyAcc-XYZ) and gravity acceleration signals (tGravityAcc-XYZ) using another low pass Butterworth filter with a corner frequency of 0.3 Hz. 

Subsequently, the body linear acceleration and angular velocity were derived in time to obtain Jerk signals (tBodyAccJerk-XYZ and tBodyGyroJerk-XYZ). 

Also the magnitude of these three-dimensional signals were calculated using the Euclidean norm (tBodyAccMag, tGravityAccMag, tBodyAccJerkMag, tBodyGyroMag, tBodyGyroJerkMag). 

Finally a Fast Fourier Transform (FFT) was applied to some of these signals producing fBodyAcc-XYZ, fBodyAccJerk-XYZ, fBodyGyro-XYZ, fBodyAccJerkMag, fBodyGyroMag, fBodyGyroJerkMag. (Note the 'f' to indicate frequency domain signals). 

These signals were used to estimate variables of the feature vector for each pattern: '-XYZ' is used to denote 3-axial signals in the X, Y and Z directions. The set of variables that were estimated from these signals are: 

  * mean(): Mean value
  * std(): Standard deviation
  * mad(): Median absolute deviation 
  * max(): Largest value in array
  * min(): Smallest value in array
  * sma(): Signal magnitude area
  * energy(): Energy measure. Sum of the squares divided by the number of values. 
  * iqr(): Interquartile range 
  * entropy(): Signal entropy
  * arCoeff(): Autorregresion coefficients with Burg order equal to 4
  * correlation(): correlation coefficient between two signals
  * maxInds(): index of the frequency component with largest magnitude
  * meanFreq(): Weighted average of the frequency components to obtain a mean frequency
  * skewness(): skewness of the frequency domain signal 
  * kurtosis(): kurtosis of the frequency domain signal 
  * bandsEnergy(): Energy of a frequency interval within the 64 bins of the FFT of each window.
  * angle(): Angle between to vectors.

Additional vectors obtained by averaging the signals in a signal window sample. These are used on the angle() variable:

  * gravityMean
  * tBodyAccMean
  * tBodyAccJerkMean
  * tBodyGyroMean
  * tBodyGyroJerkMean

Summing up, the raw data consists of 561 variables, and our study focuses 
on those related with the _mean_ and the _standard deviation_. A list of these variables and its description can be found in the section Variables of interest.

The complete list of variables of each feature vector and its 
description is available in the section Enumeration and description of all the variables.


## Transformations or work that you performed to clean up the data
We have done the following to clean and tidy the raw data:

  1. Create a data.table that gathers the ID number of each individual with the activity performed and the corresponding values of the measured variables. Since there is no indication, we assume that each row of each table (file) refers to the same individual. That is: 
     * the first row in subject_test.txt stands for individual 1 in the test subset.
     * The first row in y_test.txt stands for activity performed by individual 1  in the test subset. 
     * The first row in X_test.txt stands for the corresponding measures of all the  variables. 
   It is done for both test and training files. Let us call the data.frames  dfTest and dfTraining.
  2. Gather the  data.frames  dfTest and dfTraining.
  3. We extract only the measurements on the mean and standard deviation for each measurement. 
  4. The activities are coded with a number between 1 and 6. We set descriptive activity names to name the activities in the data set. 
  5. We set descriptive labels variable names. 

The resulting data (contained in a data.frame) is the tidy data.

## Variables of interest 
In this section we list and describe the variables mentioned in step 3. The number assigned to each variable refers to its order in the complete list of variables of the raw data includedin the next section Enumeration and description of all the variables.

```{r, echo=FALSE}
# description of the variables "subject" and "activity"
cat(paste("Variable number 1. Variable name: subject. Variable class: integer","\n",  
"Variable range: [1, 30].", "\n",  
"1 = walking", "\n",  
"2 = walking_upstairs", "\n",  
"3 = walking_downstairs", "\n",  
"4 = sitting", "\n",  
"5 = standing", "\n",  
"6 = laying", sep=""), fill = TRUE)

cat(paste("Variable number 2. Variable name: activity. Variable class: integer","\n",  
"Variable range: [1, 6].", "\n",  
"1 = walking", "\n",  
"2 = walking_upstairs", "\n",  
"3 = walking_downstairs", "\n",  
"4 = sitting", "\n",  
"5 = standing", "\n",  
"6 = laying", sep=""), fill = TRUE)

# get the name of the variables (these variables are also called features)
AllFeaturesAux <- read.table(file = "./UCI HAR Dataset/features.txt")
AllFeatures <- as.character(AllFeaturesAux[, 2])

myFeatures <- which(grepl("mean|std", x = as.character(AllFeatures)))



lines<-readLines(con = "./UCI HAR Dataset/features_info.txt", n = 49, ok = TRUE, warn = TRUE,
          encoding = "unknown", skipNul = FALSE)
myLines <- lines[33:49]

operations <- unlist(strsplit(myLines, split = ": "))[seq(from = 1, to = 2*length(myLines), by=2)]
operationsMeaning <- unlist(strsplit(myLines, split = ": "))[seq(from = 0, to = 2*length(myLines), by=2)]



operationsAux <- c()
for(i in 1:(length(operations)-1)){
     operationsAux <- paste(operationsAux, operations[i], "|", sep="")
}
operationsAux <-paste(operationsAux,operations[length(operations)], sep="" )


for(i in 1:length(myFeatures)){
     if(substring(as.character(AllFeatures[myFeatures[i]]), 1, 1) == "t"){
          slot1="time"
     }
     
     if(substring(as.character(AllFeatures[[myFeatures[i]]]), 1, 1) == "f"){
          slot1="frequency domain signal (after applying the  Fast Fourier Transform)"
     }
     
     if(grepl("Body", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot2="body"
     } 
     
     if(grepl("Gravity", x = as.character(AllFeatures[[myFeatures[i]]]))){slot2="gravity"}
     
     if(grepl("Acc", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot3="acceleration"
     } else {slot3="gyroscope"}
     
     if(grepl("Jerk", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot4="Jerk signals"
     } else {slot4=""}
     
     if(grepl("Mag", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot5="magnitude (calculated using the Euclidean norm)"
     } else {slot5 = ""}
     
     if(grepl("Mag", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot5="magnitude (calculated using the Euclidean norm)"
     }
     
     currentOperation <- unlist(strsplit(as.character(AllFeatures[[myFeatures[i]]]), split = "-"))[which(grepl(pattern = c("\\("), x =
                       unlist(strsplit(as.character(AllFeatures[[myFeatures[i]]]), split = "-"))
                  )
            )]
     
     slot6<-operationsMeaning[which(operations == currentOperation)]
     
     if(grepl("Mean", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot6="Mean"
     }
     
     
     slot7=c("")
     if(grepl("-X", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot7="X axis component"
     } 
     
     if(grepl("-Y", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot7="Y axis component"
     } 
     
     if(grepl("-Z", x = as.character(AllFeatures[[myFeatures[i]]]))){
          slot7="Z axis component"
     } 
     
     headMs = paste( "Variable number ", myFeatures[i]+2, ". Variable name", as.character(AllFeatures[[myFeatures[i]]]), ". Variable class: numeric ", sep="")
     bodyMs = paste( slot6, "of the", "\n", slot1, slot2, sep=" ")
     
     cat(paste( headMs, "\n", bodyMs,  slot3, slot4, slot5, slot7, "\n", sep = " "), fill = TRUE)

}
```



## Enumeration and description of all the variables

```{r, echo=FALSE}
cat(paste("Variable number 1. Variable name: subject. Variable class: integer","\n",  
"Variable range: [1, 30].", "\n",  
"1 = walking", "\n",  
"2 = walking_upstairs", "\n",  
"3 = walking_downstairs", "\n",  
"4 = sitting", "\n",  
"5 = standing", "\n",  
"6 = laying", sep=""), fill = TRUE)

cat(paste("Variable number 2. Variable name: activity. Variable class: integer","\n",  
"Variable range: [1, 6].", "\n",  
"1 = walking", "\n",  
"2 = walking_upstairs", "\n",  
"3 = walking_downstairs", "\n",  
"4 = sitting", "\n",  
"5 = standing", "\n",  
"6 = laying", sep=""), fill = TRUE)

cat("Note that the value of the following measures ranges are bounded so that lie in [-1,1].")


for(i in 1:length(AllFeatures)){
     
     if(substring(as.character(AllFeatures[i]), 1, 1) == "t"){
     slot1="time"
     }
     
     if(substring(as.character(AllFeatures[i]), 1, 1) == "f"){
     slot1="frequency domain signal (after applying the  Fast Fourier Transform)"
     }
     
     if(grepl("Body", x = as.character(AllFeatures[i]))){
     slot2="body"
     } 
     
     if(grepl("Gravity", x = as.character(AllFeatures[i]))){slot2="gravity"}
     
     if(grepl("Acc", x = as.character(AllFeatures[i]))){
     slot3="acceleration"
     } else {slot3="gyroscope"}
     
     if(grepl("Jerk", x = as.character(AllFeatures[i]))){
     slot4="Jerk signals"
     } else {slot4=""}
     
     if(grepl("Mag", x = as.character(AllFeatures[i]))){
     slot5="magnitude (calculated using the Euclidean norm)"
     } else {slot5 = ""}
     
     if(grepl("Mag", x = as.character(AllFeatures[i]))){
     slot5="magnitude (calculated using the Euclidean norm)"
     }
     
     currentOperation <- unlist(strsplit(as.character(AllFeatures[i]), split = "-"))[which(grepl(pattern = c("\\("), x =
                       unlist(strsplit(as.character(AllFeatures[i]), split = "-"))
                  )
            )]
     
     slot6<-operationsMeaning[which(operations == currentOperation)]
     
     if(grepl("Mean", x = as.character(AllFeatures[i]))){
     slot6="Mean"
     }
     
          slot7=c("")
     if(grepl("-X", x = as.character(AllFeatures[i]))){
     slot7="X axis component"
     }
     
     if(grepl("-Y", x = as.character(AllFeatures[i]))){
     slot7="Y axis component"
     }
     
     if(grepl("-Z", x = as.character(AllFeatures[i]))){
     slot7="Z axis component"
     }
     headMs = paste( "Variable number ", i+2, ". Variable name", as.character(AllFeatures[i]), ". Variable class: numeric ", sep="")
     bodyMs = paste( slot6, "of the", "\n", slot1, slot2, sep=" ")
     
     cat(paste( headMs, "\n", bodyMs, "\n", slot3, slot4, slot5, slot7, "\n", sep = " "), fill = TRUE)
}

```
